Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2020-11-07 paper/main.tex
remove unused

In this work, we propose and analyze Unbalanced Mallows Model (UMM), a
population-based probabilistic algorithm specifically designed for optimizing
black-box problems over a permutation landscape under a limited budget. There
exists a wide body of research on probabilistic models for permutations, such
as Mallows~\cite{FliVer1986}, that have already been been used in optimization
problems successfully, with the most prominent examples being estimation of
distribution algorithms
(EDAs)~\cite{CebIruMen2014eda,CebMenLoz2015mallows}. ACO may also be considered
a type of EDA, since it builds a probability distribution model from which
solutions are sampled with a bias towards the best solutions evaluated so
far. These classical EDAs  assume that function
evaluations are cheap and/or the fitness function is white-box.\MANUEL{Do EDAs really assume that the fitness function is white-box? ACO doesn't.}
%The proposed UMM drastically differs from the above algorithms.

@@

Permutations are defined as bijections of the set $[n]$ of integers onto itself. The set of all permutations of $n$ values is denoted as $S_n$ and has cardinality $n!$. We represent permutations as an ordered vector and, we say that position $i$ contains value $j$ whenever $\sigma(i)=j$. We denote permutations with Greek letters.\MANUEL{Do we use $e$ anywhere? If not, we do not need this definition ``with exception of the identity permutation denoted as $e=(1, 2, 3, \dotsc,n)$.''} We denote the composition of $\sigma$ and $\pi$ as $\sigma\pi$ and the inverse of $\sigma$ as $\sigma^{-1}$, for which the relation $\sigma\sigma^{-1}=(1, 2, 3, \dotsc,n)$ always holds.

@@

\subsubsection{Computing environment.}
Due to the computational effort required, we used several computing systems to
run experiments. All UMM experiments were either on run on Intel Xeon
\emph{Broadwell} E5-2683v4 CPUs at 2.10\,GHz or Intel Xeon \emph{Haswell} E5-2680v3
CPUs at 2.50\,GHz, both of them with 32\,GB RAM running Linux.\MANUEL{If you enter in a node and do \texttt{cat /etc/lsb-release} what does it say?} %
Experiments with
CEGO used a more powerful system: Intel Xeon \emph{Ivybridge} E5-2650v2 CPUs at
2.60\,GHz, 64\,GB RAM running CentOS Linux release 7.4.1708 (Core).

%processor : 23
%vendor_id : GenuineIntel
%cpu family : 6
%model : 63
%model name : Intel(R) Xeon(R) CPU E5-2680 v3 @ 2.50GHz
%stepping : 2
%microcode : 0x43
%cpu MHz : 1199.951
%cache size : 30720 KB
%physical id : 1
%siblings : 12
%core id : 13
%cpu cores : 12
%apicid : 58
%initial apicid : 58
%fpu : yes
%fpu_exception : yes
%cpuid level : 15
%wp : yes
%flags : fpu vme de pse tsc msr pae mce cx8 apic sep mtrr pge mca cmov pat pse36 clflush dts acpi mmx fxsr sse sse2 ss ht tm pbe syscall nx pdpe1gb rdtscp lm constant_tsc arch_perfmon pebs bts rep_good nopl xtopology nonstop_tsc aperfmperf eagerfpu pni pclmulqdq dtes64 monitor ds_cpl vmx smx est tm2 ssse3 sdbg fma cx16 xtpr pdcm pcid dca sse4_1 sse4_2 x2apic movbe popcnt tsc_deadline_timer aes xsave avx f16c rdrand lahf_lm abm epb intel_ppin ssbd ibrs ibpb stibp tpr_shadow vnmi flexpriority ept vpid fsgsbase tsc_adjust bmi1 avx2 smep bmi2 erms invpcid cqm xsaveopt cqm_llc cqm_occup_llc dtherm ida arat pln pts md_clear spec_ctrl intel_stibp flush_l1d
%bogomips : 5005.34
%clflush size : 64
%cache_alignment : 64
%address sizes : 46 bits physical, 48 bits virtual
%
%The node n015 has 32 processors. This is the info of the 32st processor:
%processor : 31
%vendor_id : GenuineIntel
%cpu family : 6
%model : 79
%model name : Intel(R) Xeon(R) CPU E5-2683 v4 @ 2.10GHz
%stepping : 1
%microcode : 0xb000022
%cpu MHz : 2442.608
%cache size : 40960 KB
%physical id : 1
%siblings : 16
%core id : 15
%cpu cores : 16
%apicid : 62
%initial apicid : 62
%fpu : yes
%fpu_exception : yes
%cpuid level : 20
%wp : yes
%flags : fpu vme de pse tsc msr pae mce cx8 apic sep mtrr pge mca cmov pat pse36 clflush dts acpi mmx fxsr sse sse2 ss ht tm pbe syscall nx pdpe1gb rdtscp lm constant_tsc arch_perfmon pebs bts rep_good nopl xtopology nonstop_tsc aperfmperf eagerfpu pni pclmulqdq dtes64 monitor ds_cpl vmx smx est tm2 ssse3 sdbg fma cx16 xtpr pdcm pcid dca sse4_1 sse4_2 x2apic movbe popcnt tsc_deadline_timer aes xsave avx f16c rdrand lahf_lm abm 3dnowprefetch epb cat_l3 cdp_l3 intel_ppin intel_pt tpr_shadow vnmi flexpriority ept vpid fsgsbase tsc_adjust bmi1 hle avx2 smep bmi2 erms invpcid rtm cqm rdt_a rdseed adx smap xsaveopt cqm_llc cqm_occup_llc cqm_mbm_total cqm_mbm_local dtherm ida arat pln pts
%bogomips : 4207.67
%clflush size : 64
%cache_alignment : 64
%address sizes : 46 bits physical, 48 bits virtual

@@

On the other hand, % \EKHINE{cambio el orden}
CEGO starts with a random sampling
of 10 solutions, with are quite poor as expected, followed by building and
updating the Kriging model based new evaluations, which leads to an impressive
improvement in less than 50 evaluations. However, progress is %more or less 
halted at this point and the model seems to have trouble generating better
solutions. Lower uniformity values (e.g., \manuel{$\phi=.5$}{0.5})\MANUEL{We never use $\phi$, not even in the plots, so less have one less symbol to worry about.} show worsening evaluations
despite the increase in solutions used to build the model. In summary, the
Kriging model clearly helps CEGO to quickly identify good solutions in very few
evaluations. However, for some reason, CEGO is not able to further improve
those solutions given more data. This may due to the model not being able to
produce accurate predictions or the underlying optimizer not being able to find
improved solutions using the model.
%\MANUEL{perhaps we should measure the
%  prediction error and see if it grows or decreases. That would answer this
%  question.}

@@

In the following paragraphs we analyze the result over real-world instances. %
Figure~\ref{fig:loplib} shows the results on the instances taken from LOPLIB.
  %% MANUEL: Ya dimos la URL arriba, así que no la repetiría aquí
  % \footnote{\url{http://grafo.etsii.urjc.es/optsicom/lolib/}}
The distribution\MANUEL{what do you mean the distribution? Is this a feature of
  the matrix $a$?} of these instances differs from the synthetic LOP instances,
and, thus, so does the landscape. However, UMM shows a very similar behavior:
there is a (close to) linear decrease \manuel{on the function evaluations of
  new solutions along iterations}{of the fitness of new solutions over each new
  evaluation}. This suggests that UMM \manuel{very good fits}{is able to model}
the fitness landscape. Regarding CEGO, we \manuel{can take similar
  conclusions}{observe a similar behavior} for the first 50-100
\manuel{iterations}{evaluations}. At this point, CEGO reaches the time limit of
3 days, so no observations can be made for longer budgets.  Overall, the
improvement from the first (random) to the last solutions is 2-3 times larger
in UMM than CEGO. %
\MANUEL{But this is a bit misleading because UMM runs for 4 times more
  evaluations, perhaps we should also measure up to the evaluation that CEGO
  reaches so we report both values?}

In the case of the PFSP, the progress of both algorithms resembles a random search, as illustrated by the results
 obtained on \texttt{rec05} (Fig.~\ref{fig:rec05}), which is the smallest PFSP
instance considered here. The lack of any apparent convergence suggests that
the models (both Bayesian and probabilistic) are not learning anything about
the fitness landscape. Results for larger PFSP corroborate these conclusions\supplement.
\manuel{Although previous studies~\citep{ZaeStoBar2014:ppsn} concluded
that other distance measures are more suited than Kendall's-$\tau$ distance for
the PFSP, the reported fitness differences between various distance metrics are
small.
Hence, we conjecture that the same behavior will be observed with other distance metrics.}{}\MANUEL{On second thought, I'm not sure we should say this, because we really have no idea and we actually hope it is not true.}

@@

In this paper, we have introduced UMM, a population-based probabilistic
algorithm based on an unbalanced Mallows model. The algorithm is designed for
black-box combinatorial problems on permutation landscapes and when the budget
of fitness evaluations is severely limited (up to $1000$ or less).  UMM 
\ekhine{}{has shown to perform statistically better or equal than CEGO, the state of the art algorithm, in the same instances as in the original paper of CEGO and some new instances. UMM}
is specially well-suited for budget-limited scenarios that are still
time-sensitive, e.g., when the overhead incurred by the optimizer should be no
more than a few minutes added to each fitness evaluation. As shown by our
results, a Bayesian optimizer such as CEGO may sometimes require more than hour
\EKHINE{me parece mas de CEGO q UMM y parece q UMM solo tiene de ventaja ir rápido. Primero destacaría lo bueno de UMM. Creo q es importante destaacar q UMM va MEJOR o IGUAL en TODOS los experimentos  y luego que addemás es más rápido}
\MANUEL{Pero UMM es peor en rec13 y nug12, y en LOP es mejor porque CEGO no llega a 400, es decir, porque UMM es más rápido. No podemos saber qué pasaría si CEGO llegara a 400. Podríamos medir lo que pasa en el momento en que matamos a CEGO y dar también ese valor entonces la diferencia en velocidad no importaría. }
per fitness evaluation. The time required to train the surrogate-model of CEGO
also sharply increases with permutation length, hence, UMM would be a
computationally feasible alternative for relatively large problem sizes.

>>>>

In this work, we propose and analyze Unbalanced Mallows Model (UMM), a
population-based probabilistic algorithm specifically designed for optimizing
black-box problems over a permutation landscape under a limited budget. There
exists a wide body of research on probabilistic models for permutations, such
as Mallows~\cite{FliVer1986}, that have already been been used in optimization
problems successfully, with the most prominent examples being estimation of
distribution algorithms
(EDAs)~\cite{CebIruMen2014eda,CebMenLoz2015mallows}. ACO may also be considered
a type of EDA, since it builds a probability distribution model from which
solutions are sampled with a bias towards the best solutions evaluated so
far. These classical EDAs  assume that function
evaluations are cheap and/or the fitness function is white-box.\MANUEL{Do EDAs really assume that the fitness function is white-box? ACO doesn't: They are hybridized with local search }
%The proposed UMM drastically differs from the above algorithms.

@@

Permutations are defined as bijections of the set $[n]$ of integers onto itself. The set of all permutations of $n$ values is denoted as $S_n$ and has cardinality $n!$. We represent permutations as an ordered vector and, we say that position $i$ contains value $j$ whenever $\sigma(i)=j$. We denote permutations with Greek letters.
% \MANUEL{Do we use $e$ anywhere? If not, we do not need this definition ``with exception of the identity permutation denoted as $e=(1, 2, 3, \dotsc,n)$.''}
We denote the composition of $\sigma$ and $\pi$ as $\sigma\pi$ and the inverse of $\sigma$ as $\sigma^{-1}$, for which the relation $\sigma\sigma^{-1}=(1, 2, 3, \dotsc,n)$ always holds.

@@

\subsubsection{Computing environment.}
Due to the computational effort required, we used several computing systems to
run experiments. All UMM experiments were either on run on Intel Xeon
\emph{Broadwell} E5-2683v4 CPUs at 2.10\,GHz or Intel Xeon \emph{Haswell} E5-2680v3
CPUs at 2.50\,GHz, both of them with 32\,GB RAM. 
 %
Experiments with
CEGO used a more powerful system: Intel Xeon \emph{Ivybridge} E5-2650v2 CPUs at
2.60\,GHz, 64\,GB RAM. % running CentOS Linux release 7 (Core).
The operating system was CentOS Linux release 7 (Core).
%processor : 23
%vendor_id : GenuineIntel
%cpu family : 6
%model : 63
%model name : Intel(R) Xeon(R) CPU E5-2680 v3 @ 2.50GHz
%stepping : 2
%microcode : 0x43
%cpu MHz : 1199.951
%cache size : 30720 KB
%physical id : 1
%siblings : 12
%core id : 13
%cpu cores : 12
%apicid : 58
%initial apicid : 58
%fpu : yes
%fpu_exception : yes
%cpuid level : 15
%wp : yes
%flags : fpu vme de pse tsc msr pae mce cx8 apic sep mtrr pge mca cmov pat pse36 clflush dts acpi mmx fxsr sse sse2 ss ht tm pbe syscall nx pdpe1gb rdtscp lm constant_tsc arch_perfmon pebs bts rep_good nopl xtopology nonstop_tsc aperfmperf eagerfpu pni pclmulqdq dtes64 monitor ds_cpl vmx smx est tm2 ssse3 sdbg fma cx16 xtpr pdcm pcid dca sse4_1 sse4_2 x2apic movbe popcnt tsc_deadline_timer aes xsave avx f16c rdrand lahf_lm abm epb intel_ppin ssbd ibrs ibpb stibp tpr_shadow vnmi flexpriority ept vpid fsgsbase tsc_adjust bmi1 avx2 smep bmi2 erms invpcid cqm xsaveopt cqm_llc cqm_occup_llc dtherm ida arat pln pts md_clear spec_ctrl intel_stibp flush_l1d
%bogomips : 5005.34
%clflush size : 64
%cache_alignment : 64
%address sizes : 46 bits physical, 48 bits virtual
%
%The node n015 has 32 processors. This is the info of the 32st processor:
%processor : 31
%vendor_id : GenuineIntel
%cpu family : 6
%model : 79
%model name : Intel(R) Xeon(R) CPU E5-2683 v4 @ 2.10GHz
%stepping : 1
%microcode : 0xb000022
%cpu MHz : 2442.608
%cache size : 40960 KB
%physical id : 1
%siblings : 16
%core id : 15
%cpu cores : 16
%apicid : 62
%initial apicid : 62
%fpu : yes
%fpu_exception : yes
%cpuid level : 20
%wp : yes
%flags : fpu vme de pse tsc msr pae mce cx8 apic sep mtrr pge mca cmov pat pse36 clflush dts acpi mmx fxsr sse sse2 ss ht tm pbe syscall nx pdpe1gb rdtscp lm constant_tsc arch_perfmon pebs bts rep_good nopl xtopology nonstop_tsc aperfmperf eagerfpu pni pclmulqdq dtes64 monitor ds_cpl vmx smx est tm2 ssse3 sdbg fma cx16 xtpr pdcm pcid dca sse4_1 sse4_2 x2apic movbe popcnt tsc_deadline_timer aes xsave avx f16c rdrand lahf_lm abm 3dnowprefetch epb cat_l3 cdp_l3 intel_ppin intel_pt tpr_shadow vnmi flexpriority ept vpid fsgsbase tsc_adjust bmi1 hle avx2 smep bmi2 erms invpcid rtm cqm rdt_a rdseed adx smap xsaveopt cqm_llc cqm_occup_llc cqm_mbm_total cqm_mbm_local dtherm ida arat pln pts
%bogomips : 4207.67
%clflush size : 64
%cache_alignment : 64
%address sizes : 46 bits physical, 48 bits virtual

@@

On the other hand, % \EKHINE{cambio el orden}
CEGO starts with a random sampling
of 10 solutions, with are quite poor as expected, followed by building and
updating the Kriging model based new evaluations, which leads to an impressive
improvement in less than 50 evaluations. However, progress is %more or less 
halted at this point and the model seems to have trouble generating better
solutions. Lower uniformity values, e.g., 0.5,
 show worsening evaluations
despite the increase in solutions used to build the model. In summary, the
Kriging model clearly helps CEGO to quickly identify good solutions in very few
evaluations. However, for some reason, CEGO is not able to further improve
those solutions given more data. This may due to the model not being able to
produce accurate predictions or the underlying optimizer not being able to find
improved solutions using the model.
%\MANUEL{perhaps we should measure the
%  prediction error and see if it grows or decreases. That would answer this
%  question.}

@@

In the following paragraphs we analyze the result over real-world instances. %
Figure~\ref{fig:loplib} shows the results on the instances taken from LOPLIB.
  %% MANUEL: Ya dimos la URL arriba, así que no la repetiría aquí
  % \footnote{\url{http://grafo.etsii.urjc.es/optsicom/lolib/}}
The type of instances from the synthetic LOP instances, and, thus, so does the
landscape. However, UMM shows a very similar behavior: there is a (close to)
linear decrease of the fitness of new solutions over each new evaluation. This
suggests that UMM is able to model the fitness landscape. Regarding CEGO, we
observe a similar behavior for the first 50-100 evaluations. At this point,
CEGO reaches the time limit of 3 days, so no observations can be made for
longer budgets.  Overall, the improvement from the first (random) to the last
solutions is 2-3 times larger in UMM than CEGO. %
\MANUEL{But this is a bit misleading because UMM runs for 4 times more
  evaluations, perhaps we should also measure up to the evaluation that CEGO
  reaches so we report both values?}

In the case of the PFSP, the progress of both algorithms resembles a random search, as illustrated by the results
 obtained on \texttt{rec05} (Fig.~\ref{fig:rec05}), which is the smallest PFSP
instance considered here. The lack of any apparent convergence suggests that
the models (both Bayesian and probabilistic) are not learning anything about
the fitness landscape. Results for larger PFSP corroborate these conclusions\supplement.
% \manuel{Although previous studies~\citep{ZaeStoBar2014:ppsn} concluded
% that other distance measures are more suited than Kendall's-$\tau$ distance for
% the PFSP, the reported fitness differences between various distance metrics are
% small.
% Hence, we conjecture that the same behavior will be observed with other distance metrics.}{}\MANUEL{On second thought, I'm not sure we should say this, because we really have no idea and we actually hope it is not true.}

@@

In this paper, we have introduced UMM, a population-based probabilistic
algorithm based on an unbalanced Mallows model. The algorithm is designed for
black-box combinatorial problems on permutation landscapes and when the budget
of fitness evaluations is severely limited (up to $1000$ or less).  UMM 
\ekhine{}{has shown to perform statistically better or equal than CEGO, the state of the art algorithm, in the same instances as in the original paper of CEGO and some new instances. UMM}
is specially well-suited for budget-limited scenarios that are still
time-sensitive, e.g., when the overhead incurred by the optimizer should be no
more than a few minutes added to each fitness evaluation. As shown by our
results, a Bayesian optimizer such as CEGO may sometimes require more than hour
\MANUEL{UMM is able to model the LOP landscape at least as well as CEGO as shown by the plots where UMM shows a linear decrease in fitness.}
\EKHINE{me parece mas de CEGO q UMM y parece q UMM solo tiene de ventaja ir rápido. Primero destacaría lo bueno de UMM. Creo q es importante destaacar q UMM va MEJOR o IGUAL en TODOS los experimentos  y luego que addemás es más rápido}
\MANUEL{Pero UMM es peor en rec13 y nug12, y en LOP es mejor porque CEGO no llega a 400, es decir, porque UMM es más rápido. No podemos saber qué pasaría si CEGO llegara a 400. Podríamos medir lo que pasa en el momento en que matamos a CEGO y dar también ese valor entonces la diferencia en velocidad no importaría. }
per fitness evaluation. The time required to train the surrogate-model of CEGO
also sharply increases with permutation length, hence, UMM would be a
computationally feasible alternative for relatively large problem sizes.
